Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-06-20 nemolite2d_perf.tex
nemolite2d-paper: update cpu-gpu comparison data. Add discussion of GPU optimisation results.

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
approach.

@@

When using the OpenACC \texttt{kernels} directive the compiler creates
GPU kernels by parallelising only the outer loop of each loop nest.
For the majority of loop nests in NEMOLite2D all of the iterations are
independent and the loop bodies consist of just a handful of
executable statements. For small kernels the loop start-up cost can be
significant and therefore it is beneficial to generate as many threads
as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
data size is greatre than the number of threads). For this approach to
be efficient we must expose the maximum amount of parallelism to the
GPU and we do this by instructing the compiler to parallelise both
levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
achieved using the OpenACC \texttt{loop collapse(2)} directive for all
of the smaller kernels in NEMOLite2D.

@@

In contrast to all of the other NEMOLite2D kernels, the Momentum
kernels ($u$ and $v$) are relatively large in terms of the number of
executable statements they contain. In turn, this means that the
corresponding kernels will require more state and thus more registers
on the GPU. Therefore, rather than generating as many threads as will
fit on the GPU, it is more efficient only to generate as many as
required by the problem so as to minimise reigster pressure. Once
slots become free on the GPU, new threads will launch with the
associated cost amortized by the longer execution time of the larger
kernel. The compiler can be persuaded
to adopt the above strategy through the use of the (PGI-specific)
\texttt{loop shortloop} directive which tells it that there is not
likely to be much thread re-use in the following loop. This directive
was applied to both the \texttt{ji} and \texttt{jj} loops in both of
the Momentum kernels.

@@

Both of the Momentum kernels read from 20 double-precision arrays and
thus require considerable memory bandwidth.  However, the majority of
these arrays are used by both the $u$ and $v$ kernels. It is therefore
possible to reduce the memory-bandwidth requirements by fusing the two
kernels together. In practice, this means fusing the two,
doubly-nested loops so that we have a single loop nest containing both
the $u$ and $v$ kernel bodies/calls.

@@

We also experimented with using CUDA directly within the original form
of NEMOLite2D in order to quantify any performance penalty incurred by
taking the {PS}y{KA}l/OpenACC approach.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{opt_stages_256}
\caption{Performance of the {PS}y{KA}l version of NEMOLite2D for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  NEMOLite2D for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.

The performance of these two versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring of
NEMOLite2D did not prove to be a barrier to achieving a performant
code when using OpenACC directives, we only report the performance of
the {PS}y{KA}l OpenACC version. For problem sizes up to $256^2$ the
difference in the performance of the two versions is less than 5\% and
for the larger domains it is at most 11\%. This demonstrates that the
performance cost of the {PS}y{KA}l approach in utilising a GPU for
NEMOLite2D is minimal.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the {PS}y{KA}l (OpenACC) and CUDA GPU
  implementations of NEMOLite2D. The performance for the OpenACC
  version of the original code is not shown since it is identical to
  that of the {PS}y{KA}l version.}
\label{FIG_gpu_opt_stages}
\end{figure}

>>>>

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. OpenACC
directives were used as that approach is similar to the use of OpenMP
directives and works well within the {PS}y{KA}l approach. In order to
quantify any performance penalty incurred by taking the
{PS}y{KA}l/OpenACC approach, we experimented with using CUDA directly
within the original form of NEMOLite2D.

@@

When using the OpenACC \texttt{kernels} directive the compiler creates
GPU kernels by parallelising only the outer loop of each loop nest.
For the majority of loop nests in NEMOLite2D all of the iterations are
independent and the loop bodies consist of just a handful of
executable statements. For small kernels the loop start-up cost can be
significant and therefore it is beneficial to generate as many threads
as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
data size is greater than the number of threads). For this approach to
be efficient we must expose the maximum amount of parallelism to the
GPU and we do this by instructing the compiler to parallelise both
levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
achieved using the OpenACC \texttt{loop collapse(2)} directive for all
of the smaller kernels in NEMOLite2D.

@@

In contrast to all of the other NEMOLite2D kernels, the Momentum
kernels ($u$ and $v$) are relatively large in terms of the number of
executable statements they contain. In turn, this means that the
corresponding kernels will require more state and thus more registers
on the GPU. Therefore, rather than generating as many threads as will
fit on the GPU, it is more efficient only to generate as many as
required by the problem so as to minimise register pressure. Once
slots become free on the GPU, new threads will launch with the
associated cost amortized by the longer execution time of the larger
kernel. The compiler can be persuaded
to adopt the above strategy through the use of the (PGI-specific)
\texttt{loop shortloop} directive which tells it that there is not
likely to be much thread re-use in the following loop. This directive
was applied to both the \texttt{ji} and \texttt{jj} loops in both of
the Momentum kernels.

@@

Both of the Momentum kernels read from 16 double-precision arrays and
thus require considerable memory bandwidth.  However, the majority of
these arrays are used by both the $u$ and $v$ kernels. It is therefore
possible to reduce the memory-bandwidth requirements by fusing the two
kernels together. In practice, this means fusing the two,
doubly-nested loops so that we have a single loop nest containing both
the $u$ and $v$ kernel bodies/calls.

@@

We also experimented with using CUDA directly within the original form
of NEMOLite2D in order to quantify any performance penalty incurred by
taking the {PS}y{KA}l/OpenACC approach. To do this we used PGI's
support for CUDA Fortran to create a CUDA kernel for the (fused)
Momentum kernel. The only significant code change with this approach
is the explicit set-up of the grid- and block-sizes and the way in
which the kernel is launched (i.e. use of the \texttt{call kernel $<<<$
  gridDim1, blockDim1 $>>>$(args)} syntax).

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\picwidth]{opt_stages_256}
\caption{Serial performance of the {PS}y{KA}l version of NEMOLite2D for the
  $256^{2}$ domain at each stage of optimisation. The first (black)
  bar of each cluster gives the performance of the original version of
  NEMOLite2D for that compiler/CPU combination.}
\label{FIG_opt_stages_256}
\end{figure}

@@

In contrast to the CPU, we have just a single compiler to consider
when looking at OpenACC performance. We are therefore able to
investigate the effect of the various optimisations described in
Section~\ref{sec_oacc_steps} for the full range of problem sizes.  All
of the reported performance figures were obtained on an NVIDIA Tesla
K40m GPU running in `boost' mode (enabled with the command
\texttt{nvidia-smi -ac 3004,875}).

The performance of the various versions of the code is plotted in
Figure~\ref{FIG_gpu_opt_stages}.  Since the {PS}y{KA}l restructuring
of NEMOLite2D did not prove to be a barrier to achieving a performant
code when using OpenACC directives, we only report the performance of
the {PS}y{KA}l OpenACC versions.

\begin{figure}
\centering
\includegraphics[width=\picwidth]{gpu_opt_stages}
\caption{Performance of the {PS}y{KA}l (OpenACC) and CUDA GPU
  implementations of NEMOLite2D. Optimisations are added to the code
  in an incremental fashion. The performance for the OpenACC version
  of the original code is not shown since it is identical to that of
  the Vanilla {PS}y{KA}l version.}
\label{FIG_gpu_opt_stages}
\end{figure}

The smallest domain ($64^2$) does not contain sufficient parallelism
to fully utilise the GPU and only the parallelisation of the Flather
kernels has much effect on performance. In fact, this is a significnt
optimisation for all except the largest domain size where only changes
to the Momentum kernel yield significant gains. Collapsing the 2D loop
nests has a small but beneficial effect for the majority of problem
sizes. Gains here are limited by the fact that this optimisation was
only applied to the non-Momentum kernels which only account for som
30\% of the run-time.

Given the significance of the two Momentum kernels in the execution
profile it is no surprise that their optimisation yields the most
significant gains. Using the \texttt{shortloop} directive gives
roughly a 25\% increase in performance for domains of $512^2$ and
greater. Fusing the two momentum kernels gives a further boost of
around 5\%.

Note that all of the previous optimisations are restricted to the PSy
layer and in most cases are simply a case of adding directives or
clauses to directives. However, the question then arises as to the
cost of restricting optimisations to the PSy layer. In particular, how
does the performance of the OpenACC version of NEMOLite2D compare with
a version where those restrictions are lifted?
Figure~\ref{FIG_gpu_opt_stages} also shows the performance of the
version of the code where the (fused) Momentum kernel has been
replaced by a CUDA kernel. For domains up to $256^2$ the difference in
the performance of the two versions is less than 5\% and for the
larger domains it is at most 12\%. This demonstrates that the
performance cost of the {PS}y{KA}l approach in utilising a GPU for
NEMOLite2D is minimal. Discussions with NVIDIA regarding the
discrepancy in performance between the CUDA and OpenACC versions are
also on-going.
